\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage{a4wide}
\usepackage{amsmath,amssymb}
\usepackage{color}
\usepackage{boxproof}
\usepackage{daymonthyear}
\usepackage{float} % for the H placement identifier
\usepackage{listings}
\usepackage{stmaryrd}

%========== DEFINITIONS ==========%

\definecolor{c_comment}{rgb}	{0.38, 0.62, 0.38}
\definecolor{c_keyword}{rgb}	{0.10, 0.10, 0.81}
\definecolor{c_identifier}{rgb}	{0.00, 0.00, 0.00}
\definecolor{c_string}{rgb}		{0.50, 0.50, 0.50}

%===== SETTINGS =====%

\def\meta#1{\mbox{$\langle\hbox{#1}\rangle$}}
\def\macrowitharg#1#2{{\tt\string#1\bra\meta{#2}\ket}}

{\escapechar-1 \xdef\bra{\string\{}\xdef\ket{\string\}}}

\def\intro#1{{#1}{\cal I}}
\def\elim#1{{#1}{\cal E}}

\showboxbreadth 999
\showboxdepth 999
\tracingoutput 1

\let\imp\to
\def\elim#1{{{#1}{\cal E}}}
\def\intro#1{{{#1}{\cal I}}}
\def\lt{<}
\def\eqdef{\overset{\mathrm{def}}{=\joinrel=}}
\def\eps{\mathrel{\epsilon}}
\def\biimplies{\leftrightarrow}
\def\flt#1{\mathrel{{#1}^\flat}}
\def\setof#1{{\left\{{#1}\right\}}}
\let\implies\to
\def\KK{{\mathsf K}}
\let\squashmuskip\relax

\lstset
{
	numbers=left,
	frame=single,
	basicstyle=\footnotesize\ttfamily,
	tabsize=4,
	% colors
	commentstyle=\color{c_comment},
	keywordstyle=\color{c_keyword},
	identifierstyle=\color{c_identifier},
	stringstyle=\color{c_string},
}

%===== META DATA =====%

\title
{
{\Large Logic in Computer Science}\\
Assignment 5
}
\author
{
	Casper B. Hansen\\
	University of Copenhagen\\
	Department of Computer Science\\
	{\tt fvx507@alumni.ku.dk}
}
\date{\today}

%===== DOCUMENT =====%

\begin{document}

\maketitle

\section*{3.4.7 \mdseries Let ${\cal M} = (S,\imp,L)$ be any model for CTL and
let $\llbracket\phi\rrbracket$ denote the set of all $s \in S$ such that
${\cal M}, s \models \phi$. Prove the following set identities by inspecting
the clauses of Definition 3.15 from page 211.}
% 3.4: 7 (d-f)

\subsection*{(d) \mdseries $\llbracket\phi_1 \land \phi_2\rrbracket =
\llbracket\phi_1\rrbracket\cap\llbracket\phi_2\rrbracket$}
% 4.	M, s |= φ1 ∧ φ2 iff. M, s |= φ1 and M, s |= φ2
We know, by clause $4$ that ${\cal M}, s \models \phi_1 \land \phi_2$ iff
${\cal M}, s \models \phi_1$ and ${\cal M}, s \models \phi_2$. That is, the
set of states for which $\phi_1 \land \phi_2$ holds must then contain both of
the latter, but only states for which both of $\llbracket\phi_1\rrbracket$ and
$\llbracket\phi_2\rrbracket$ hold. Hence
$\llbracket\phi_1 \land \phi_2\rrbracket$ is, by definition, the intersection
of $\llbracket\phi_1\rrbracket$ and $\llbracket\phi_2\rrbracket$, and thereby
$\llbracket\phi_1 \land \phi_2\rrbracket =
\llbracket\phi_1\rrbracket\cap\llbracket\phi_2\rrbracket$.

\subsection*{(e) \mdseries $\llbracket\phi_1 \lor \phi_2\rrbracket =
\llbracket\phi_1\rrbracket\cup\llbracket\phi_2\rrbracket$}
% 5.	M, s | = φ1 ∨ φ2 iff. M, s |= φ1 or M, s |= φ2
By clause $5$ we have that ${\cal M}, s \models \phi_1$ or ${\cal M}, s
\models \phi_2$. That is, any state contained within either of
$\llbracket\phi_1\rrbracket$ or $\llbracket\phi_2\rrbracket$ satisfies the
model. So, $\llbracket\phi_1\rrbracket$ and $\llbracket\phi_2\rrbracket$ are
both subsets of satisfactory states. Since no other states satisfies the
model, we can rightfully conclude that $\llbracket\phi_1 \lor \phi_2\rrbracket
= \llbracket\phi_1\rrbracket\cup\llbracket\phi_2\rrbracket$.

\subsection*{(f) \mdseries $\llbracket\phi_1 \imp \phi_2\rrbracket =
(S - \llbracket\phi_1\rrbracket)\cup\llbracket\phi_2\rrbracket$}
% 6.	M, s |= φ1 → φ2 iff. M, s |≠ φ1 or M, s |= φ2
We see that by clause $6$ we have that ${\cal M}, s \not\models \phi_1$ or
${\cal M}, s \models \phi_2$. The former gives us $S -
\llbracket\phi_1\rrbracket$, as this signifies the set of states for which
$\phi_1$ does not hold. The latter gives us simply $\llbracket\phi_2
\rrbracket$. And thus, since just one of these expressions must hold, forming
the union of these --- by the same {\it or}-argument as above --- gives us
exactly that $\llbracket\phi_1 \imp \phi_2\rrbracket =
(S - \llbracket\phi_1\rrbracket)\cup\llbracket\phi_2\rrbracket$.

\newpage
\section*{3.4.10 \mdseries Which of the following pairs of CTL formulas are
equivalent? For those which are not, exhibit a model of one of the pair which
is not a model of the other}
% 3.4: 10 (a-c)

\subsection*{(a) \mdseries $\text{EF}{\ }\phi$ and $\text{EG}{\ }\phi$}
For a model ${\cal M} = (S,\imp,L)$ consisting of the transitions $s \imp t$,
$s \imp u$ and $t \imp u$, in which only $\phi$ only holds for $t$. We see
that ${\cal M}, s \models \text{EF}{\ }\phi$ holds, as we can get from the
start node $s$ to $t$ for which $\phi$ holds. However, from $t$ we could go
to $u$ for which $\phi$ doesn't hold, hence
${\cal M}, s \not\models \text{EG}{\ }\phi$.

\subsection*{(b) \mdseries $\text{EF}{\ }\phi \lor \text{EF}{\ }\psi$ and
$\text{EF}{\ }(\phi \lor \psi)$}
Under the assumption that $\text{EF}{\ }\phi \lor \text{EF}{\ }\psi$ holds, it
then follows that there exists a future state for which either $\phi$ or
$\psi$ holds. That is, by clause $5$ definition 3.15, we have that
${\cal M}, s \models \text{EF}{\ }\phi$ or
${\cal M}, s \models \text{EF}{\ }\psi$. By expanding $\phi$ of clause $12$
for ${\cal M}, s \models \text{EF}{\ }(\phi \lor \psi)$ we get just that;
$\phi \lor \psi$ must hold at some point in the future, which is satisfied.

\subsection*{(c) \mdseries $\text{AF}{\ }\phi \lor \text{AF}{\ }\psi$ and
$\text{AF}{\ }(\phi \lor \psi)$}
Consider a model ${\cal M}, s \models (S,\imp,L)$ which includes three states
$s$, $t$ and $u$, for which $s$ none of $\phi$ or $\psi$ hold, that is $L(s)
= \emptyset$. For $t$ and $u$, $\phi$ and $\psi$ holds respectively, that is
$L(t) = \phi$ and $L(u) = \psi$. Given the transitions $s \imp t$ and $s \imp
u$ we can confirm that ${\cal M}, s \models AF{\ }\phi \lor AF{\ }\psi$, as
well as ${\cal M}, s \models AF{\ }(\phi \lor \psi)$. However, if we add the
transitions $t \imp t$ and $u \imp u$, we see that we can go to either of
these, and stay there forever showing that
${\cal M}, s \not\models AF{\ }\phi \lor AF{\ }\psi$.

\newpage
\section*{NuSMV \mdseries Attached assignment}
% attached NuSMV assignment.
\subsection*{Model}
\lstinputlisting[basicstyle=\ttfamily\tiny]{mutex.smv}

\subsection*{Specifications}
I have specified safety by the condition that $\text{AG} \neg( (p_0.state = c)
\land (p_1.state = c) )$, meaning that no where in the execution of the program
will a process enter its critical section if the other is.

As for liveness I have that $\text{AG} ( (p_i.state = t) \imp \text{AF}
(p_i.state = c) )$, meaning that for any process in its trying state will
eventually reach its critical state.

\newpage
\subsection*{Safety trace}
\lstinputlisting{safety-trace.out}
As lines $42$ and $52$ of the trace shows, this did in fact occur; both
processes entered their critical states.

What happens is that $p_1$ starts running, while $p_0$ has yet to flag its
desire to enter its critical section. The $p_1$ process runs to line $3$ and
$p_0$ enters its trying state. At this point $p_1$ has yet to execute line $3$
and as such $p_0$ is free to jump directly over the {\tt while}-loop, as it
believes it is its turn. So $p_0$ moves on to line $4$, effectively entering
its critical section. Now, $p_1$ executes line $3$, making it its turn, and so
it checks the {\tt while} condition, and jumps to line $4$, also entering its
critical section. And thus, we have a violation of {\it safety}.

\subsection*{Liveness trace}
\lstinputlisting{liveness-trace.out}
Examining the trace output, we see that on line $40$ the process $p_0$ enters
its trying state $t$, but did not reach its critical state as a consequence of
the looped state path as of line $36$.

What happens here is that $p_1$ starts running, entering its trying state by
flagging its desire to enter its critical section on line $0$. It continues
onwards to line $3$, as $turn \neq 1$ and $flag[0]$ is false. Here $p_0$
starts running, entering its trying state on line $0$. Back in $p_1$ $turn$
gets set to $1$, and $p_1$ loops back to check whether or not it is its turn.

From here on out, we get into a loop of states; $p_0$ moves on to line $2$,
and then $p_1$ moves on to line $4$ as it is its turn. Moving on, $p_1$ goes
on to enter its critical section on line $5$, and on line $6$ it resets its
flag to false, and loops back to line $0$, while $p_0$ goes to line $1$
because $flag[1]$ is false. This produces a loop of states for which $p_0$
never reaches its critical state, even though it did flag its desire to do so.

The case of $p_1$ never reaching its critical state is symmetric to the above.

\end{document}